%!TEX root = main.tex

\smallskip\noindent
{\bf Benchmarks}. To extensively evaluate the OCaml LoFIRRTL Interpreter,
we mainly collect LoFIRRTL circuits from the Chisel/Firrtl execution engine--Treadle\footnote{\url{https://github.com/chipsalliance/treadle}.} and the Chisel Book--Digital Design with Chisel~\cite{chisel:book},
because Treadle is more likely to execute correctly on them so that
the conformance testing of the formalized operational semantics has high confidence.

There are 55 LoFIRRTL circuits inside the ``chipsalliance/treadle/samples'' directory
of Treadle, among which 18 LoFIRRTL circuits are excluded because they are not self-contained (i.e., dependent upon external modules),
or contain undefined semantics for memories, or have no output ports or cannot be handled by Treadle due to cyclical connection of ports.
Thus, we consider the remaining 37 LoFIRRTL circuits from Treadle, among which 3 LoFIRRTL circuits have some \prettyll{inst of} statements,
i.e., hierarchies of the modules.

There are 51 Chisel examples in Scala inside the ``schoeberl/chisel-book/src/main/scala'' directory of the Chisel Book.
These 51 Chisel examples are transformed into 77 LoFIRRTL circuits by applying the convertor \textsf{ChiselStage} of the Chisel compiler framework
to each subclass of the class $\tt Module$ in Scala. (Note that some Chisel examples contain multiple such subclasses, thus yield multiple LoFIRRTL circuits.)
Among 77 LoFIRRTL circuits, there are 67 \prettyll{inst of}-free LoFIRRTL circuits and 10 none-\prettyll{inst of}-free LoFIRRTL circuits.

%For those from Treadle are 55 LoFIRRTL circuits inside the ``chipsalliance/treadle/samples'' directory.
%We consider those LoFIRRTL circuits because Treadle is more likely to execute correctly on them so that
%the correctness testing of the formalized operational semantics has high confidence.
%Among 55 LoFIRRTL circuits, we exclude 18 LoFIRRTL circuits because they violate the syntax of
%LoFIRRTL (i.e., dependency of external modules), undefined semantics for memories, have no output ports or
%cannot be handled by Treadle due to cyclical connection of ports.
%Thus, we collected 37 LoFIRRTL circuits from Treadle, among them 3 circuits have instof statement,  which means a nested module structure.
%Beside, we generated 76 LoFIRRTL circuits from book project of Digital Design with Chisel, which containing 51 Chisel examples inside the ``schoeberl/chisel-book/src/main/scala'' directory. The transformation is executed by Chisel stage for each class extends from class $\tt Module$, as it is legal to define multi-class in one Scala program, it is possible to execute 76 LoFIRRTL circuits from 51 Chisel programs. The generated LoFIRRTL circuits including 66 single module circuits and 10 with instance structure.
Besides the above LoFIRRTL circuits, we also collect all the four LoFIRRTL circuits from
\cite{HarrisonHA21} which have been tested using a formal Agda-based Interpreter designed for a subset of LoFIRRTL.
and design three additional LoFIRRTL circuits in consideration of the coverage of all semantic rules and different hierarchical structures of the modules.

In summary, we use 121 (37+77+4+3) LoFIRRTL circuits all of which can be handled by Treadle.

\smallskip\noindent
{\bf Testing method}.
Since the input LoFIRRTL circuits to our LoFIRRTL OCaml Interpreter should be in the pseudo-SSA form
while all the collected LoFIRRTL circuits are not necessary in the pseudo-SSA form,
to resolve this issue, we implement a preprocessor to transform LoFIRRTL circuits into their pseudo-SSA based on our verified
topological sorting algorithm.
Furthermore, LoFIRRTL circuits themselves require input signals to execute.
For this purpose, we also implement a deriver in OCaml which first automatically identifies
all the input ports and their data types for each LoFIRRTL circuit and then randomly generates 100 test input signals for
the LoFIRRTL circuit. Finally, for each given LoFIRRTL circuit and each its test input,
we compare the equivalence of output ports produced by our LoFIRRTL OCaml Interpreter and Treadle.
We do not compare with the Agda-based Interpreter because it cannot directly execute circuits in LoFIRRTL.

\begin{table}[t]
\caption{Results of conformance testing}\footnotesize
\label{tab:conformancetesting}
\scalebox{0.9}{\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
\multicolumn{2}{|c|}{\multirow{2}{*}{\bf Benchmark name}} & \multirow{2}{*}{\bf \#LOC} & \multicolumn{3}{c|}{\bf Execution time (ms)} & \multirow{2}{*}{\bf Result} \\ \cline{4-6}
 \multicolumn{2}{|c|}{}                                   &                        & {\bf Treadle}  &  {\bf Ours}  & {\bf Speedup}    &  \\ \hline
\multirow{34}{*}{\textbf{\rotatebox{270}{Treadle}}} & \textbf{AggregatePassThrough}              & 72  & 0.338 & 0.074 & 3.590  & \checkmark \\
                                   & \textbf{ALU}                               & 61  & 0.905 & 0.125 & 6.219  & \checkmark \\
                                   & \textbf{BundlePassThrough}                 & 27  & 0.234 & 0.020 & 10.492 & \checkmark \\
                                   & \textbf{ConstantTapTransposedStreamingFIR} & 35  & 0.280 & 0.046 & 5.023  & \checkmark \\
                                   & \textbf{DoesAbs}                           & 36  & 0.464 & 0.047 & 8.822  & \checkmark \\
                                   & \textbf{DriverTest}                        & 8   & 0.193 & 0.002 & 97.851 & \checkmark \\
                                   & \textbf{EqModule}                          & 9   & 0.191 & 0.008 & 23.763 & \checkmark \\
                                   & \textbf{ExampleModule}                     & 44  & 0.236 & 0.059 & 3.016  & \checkmark \\
                                   & \textbf{FixedPointAdder}                   & 11  & 0.230 & 0.007 & 30.161 & \checkmark \\
                                   & \textbf{FixedPointDivide}                  & 10  & 0.193 & 0.007 & 28.099 & \checkmark \\
                                   & \textbf{FixedPointReduce}                  & 43  & 0.386 & 0.110 & 2.508  & \checkmark \\
                                   & \textbf{FixedPointShifter}                 & 15  & 0.239 & 0.012 & 18.790 & \checkmark \\
                                   & \textbf{FixedPrecisionChanger}             & 10  & 0.276 & 0.005 & 51.454 & \checkmark \\
                                   & \textbf{FixedRing1}                        & 26  & 0.264 & 0.045 & 4.907  & \checkmark \\
                                   & \textbf{GCD}                               & 30  & 0.274 & 0.043 & 5.436  & \checkmark \\
                                   & \textbf{HardwareHashMaker}                 & 122 & 0.459 & 2.894 & -0.841 & \checkmark \\
                                   & \textbf{IntegerModule}                     & 11  & 0.166 & 0.029 & 4.650  & \checkmark \\
                                   & \textbf{OverflowTypeCircuit}               & 34  & 0.243 & 0.033 & 6.325  & \checkmark \\
                                   & \textbf{ParameterizedNumberOperation}      & 12  & 0.237 & 0.025 & 8.576  & \checkmark \\
                                   & \textbf{PeekPokeBundleSpecMyCircuit}       & 17  & 0.253 & 0.009 & 28.056 & \checkmark \\
                                   & \textbf{RealGCD2}                          & 44  & 0.274 & 0.066 & 3.179  & \checkmark \\
                                   & \textbf{RealGCD3}                          & 48  & 0.441 & 0.073 & 5.053  & \checkmark \\
                                   & \textbf{RemoveMantissa}                    & 10  & 0.206 & 0.004 & 45.697 & \checkmark \\
                                   & \textbf{RingModule}                        & 11  & 0.192 & 0.018 & 9.644  & \checkmark \\
                                   & \textbf{ShouldBeBadUIntSubtractWithGrow}   & 14  & 0.258 & 0.010 & 23.763 & \checkmark \\
                                   & \textbf{SignedAdder}                       & 11  & 0.245 & 0.011 & 21.989 & \checkmark \\
                                   & \textbf{SignedModule}                      & 34  & 0.156 & 0.034 & 3.627  & \checkmark \\
                                   & \textbf{SimpleAdder}                       & 15  & 0.219 & 0.010 & 20.500 & \checkmark \\
                                   & \textbf{SimpleCaseClassModule}             & 10  & 0.226 & 0.002 & 98.158 & \checkmark \\
                                   & \textbf{SimpleComplexAdder}                & 40  & 0.295 & 0.058 & 4.083  & \checkmark \\
                                   & \textbf{Sort}                              & 85  & 0.571 & 0.268 & 1.135  & \checkmark \\
                                   & \textbf{UseMyBundle}                       & 59  & 0.261 & 0.044 & 4.868  & \checkmark \\
                                   & \textbf{VecPassThrough}                    & 75  & 0.411 & 0.067 & 5.142  & \checkmark \\
                                   & \textbf{VF}                                & 64  & 0.195 & 0.070 & 1.797  & \checkmark \\ \hline
\multirow{4}{*}{\textbf{\cite{HarrisonHA21}}}      & \textbf{Accumulator}                       & 13  & 0.175 & 0.007 & 25.705 & \checkmark \\
                                   & \textbf{Combinational}                     & 12  & 0.167 & 0.007 & 21.328 & \checkmark \\
                                   & \textbf{FibCircuit}                        & 14  & 0.190 & 0.006 & 29.195 & \checkmark \\
                                   & \textbf{ShiftRegister}                     & 21  & 0.190 & 0.010 & 18.433 & \checkmark  \\ \hline
     & \textbf{mytest}             & 29  & 0.243 & 0.029 & 7.379 & \checkmark \\  \hline
\end{tabular}}
\end{table}
%

\smallskip\noindent
{\bf Testing results}.
The results of conformance testing are reported in Table~\ref{tab:conformancetesting}, where
column (\#LOC) shows the number of locations,
column (Treadle) shows the average execution time $t_1$ of Treadle in millisecond,
column (Ours) shows the average execution time $t_2$ of our LoFIRRTL OCaml Interpreter in millisecond,
column (Speedup) is calculated via $\frac{t_1-t_2}{t_2}$,
column (Result) shows the comparison result of output ports between Treadle and our LoFIRRTL OCaml Interpreter with
\checkmark indicating the same values of the output ports under all the 100 input tests.

We can observe that our LoFIRRTL OCaml Interpreter produces the same results as Treadle on all the 37 benchmarks under all the input tests.
Furthermore, our LoFIRRTL OCaml Interpreter is significantly more efficient than Treadle on almost all the benchmarks except for HardwareHashMaker.
After an in-depth analysis, we found that HardwareHashMaker has 20 64-bit multiplication operations significantly more than the other benchmarks,
our LoFIRRTL OCaml Interpreter is extracted from the bit-wise multiplication in Coq while Treadle uses (bit) integer multiplication.
Though we could adopt (bit) integer multiplication as well,
we chose bit-wise multiplication based on the bit-vector library in Coq whose correctness has been formally proved. %and the ocaml bit-vector operations are extracted from coq code


In order to have some measure of the effectiveness of our testing, we analyzed the branch coverage of our LoFIRRTL OCaml Interpreter.
All the 38 benchmarks together cover 36 branches out of 45. To cover the remaining branches, we implement an additional benchmark, called mytest,
on which Treadle and our LoFIRRTL OCaml Interpreter also produce the same results.
 
