%!TEX root = main.tex

In this section, we show how the lowering transformations of HiFIRRTL, which were described informally in Section~\ref{subsec-lowering-transformations}, are formalized in Coq. Specifically, we consider the formalization of the six lowering transformations, namely, ``InferTypes'', ``InferWidths'', ``InferResets'', ``ExpandConnects'', ``ExpandWhens'', and ``LowerTypes''.

For simplicity, let us focus on a HiFIRRTL circuit $A$ that contains no ``inst of'' statements, that is, circuit $A$ contains exactly one module $A$.

The formalization utilizes a structure called \emph{circuit diagrams} defined below.   

\begin{definition}[Circuit diagrams]\label{def-cd}
A \emph{circuit diagram} $D$ is a triple $(ss, ce, cm)$, where $ss$ is a sequence of statements, $ce$ is a component environment that maps each component to its type, and $cm$ is a connect mapping that maps each component to an expression that is connected to it. 
\end{definition}
Note that the types in the range of $ce$ in Definition~\ref{def-cd} are not necessarily ground types, that is, they are aggregate types in general, since HiFIRRTL may contain high-level constructs, e.g. vectors and bundles. Intuitively, $ss$ is the sequence of statements in a HiFIRRTL program, while $ce$ and $cm$ extract some necessary information from $ss$ to facilitate the formalization of the lowering transformations. Specifically, $ce$ extracts the type information from $ss$ and $cm$ extracts for each component the expression that is connected to the component according to the last-connect semantics. As we will see later on, $ss$ will be updated by ``ExpandConnects'', ``ExpandWhens'', and ``LowerTypes'', and $ce$ will be modified by ``InferTypes'', ``InferWidths'', ``InferResets'',  and ``LowerTypes'',  while $cm$ will only be changed by ``ExpandWhens'', since it is the only pass that deals with the last-connect semantics. 

Before applying the lowering transformations, the module $A$ will be first transformed into a circuit diagram $D_A = (ss_0, ce_0, cm_0)$, where $ss_0$ is the sequence of statements in $A$, $ce_0$ is the component environment obtained from the component declarations in $A$, possibly with the types of some components missing or partial, and $cm_0$ is the trivial mapping that maps each component to $\bot$. (Intuitively, this means that none of the components have been connected to so far.) The missing or partial types in $ce_0$ will be completed by the lowering transformations ``InferTypes'' and ``InferWidths''.

In the sequel, we utilize the concept of circuit diagrams to illustrate how the six lowering transformations are formalized in Coq. 

%A HiFIRRTL module $M$ will be first transformed into a circuit diagram $D_M = (ss, ce, cm)$, where $ss$ is the sequence of statements in $M$, $ce$ is the component environment obtained from the variable declarations in $M$, possibly with the types of some variables missing or partial, and $cm$ is the trivial mapping that maps each component to $\bot$ (Intuitively, this means that the connects of the components have not be determined yet). The missing or partial  types in $ce$ will be completed by the compilation passes during the transformation of $M$ into a LoFIRRTL module. Moreover, the connect mapping $cm$ will be updated by the ExpandWhens compilation pass. Furthermore, the statement sequence $ss$ will be updated by ExpandConnects, ExpandWhens, LowerTypes, and RemoveReset, while they keep unchanged during the other compilation passes.  

%A description of the formalization of InferTypes, InferWidths, InferResets, ExpandConnects, ExpandWhens, LowerTypes, and RemoveResets. 

\subsubsection{Formalization of InferTypes}
Initially, the component environment $ce$ of a circuit diagram $D_M$
is empty.  The Infer Types pass is then applied to fill $ce$ with
all the components that are defined in the statement sequence $ss$,
along with their component kinds and data types.
For each component
defined in $ss$, the {\tt infer\_type} function will update $ce$
accordingly. For example, if there is a \prettyll{wire} statement,
it is added to component environment by
  \begin{gather}
    \infer[]
    {\seqq{ce \vdash {\tt infer\_type} (\mbox{wire}\;v\,:\;t_v) \rightsquigarrow ce'}}
    {\seqq{ce':= ce \uplus \{v : (t_v, Wire)\}}}.
    \end{gather}
It is the same for the \prettyll{reg} statement and ports declarations, where types are defined directly.
If it is a node-declaration statement, an entry for the node is added
to $ce$ with its type calculated by the expression $e$
assigned to the node.
\begin{gather}
    \inferii[]
    {\seqq{ce \vdash {\tt infer\_type} (\mbox{node}\;v\,=\;e) \rightsquigarrow ce'}}
    {\seqq{ce \vdash t_e:={\tt type\_of\_expr}(e)}} {\seqq{ce':= ce \uplus \{v : (t_e, Node)\}}}.
    \end{gather}
The connect statements will not introduce new components. However,
they can be used to check whether the connection is made between expressions of
equivalent types. By type equivalence, we mean that the structures of the types are the same,
but the widths are not necessarily the same.
If the types in the conect statements are equivalent,
the components environment is kept.
\begin{gather}
    \inferiii[]
    {\seqq{ce \vdash {\tt infer\_type} (v<=e) \rightsquigarrow ce}}
    {\seqq{ce \vdash v:t_v}}
    {\seqq{ce \vdash t_e:={\tt type\_of\_expr}(e)}} {\seqq{{\tt equiv}(t_v,t_e)}}
    \end{gather}
If the type equivalence is violated, an error will be raised.
\begin{gather}
    \inferiii[]
    {\seqq{ce \vdash {\tt infer\_type} (v<=e) \rightsquigarrow \bot}}
    {\seqq{ce \vdash v:t_v}}
    {\seqq{ce \vdash t_e:={\tt type\_of\_expr}(e)}} {\seqq{\neg{\tt equiv}(t_v,t_e)}}
    \end{gather}
After applying the Infer Types pass, types with an
implicitly defined width are assigned with 0 in $ce$.

\subsubsection{Formalization of InferWidths}
It is allowed to leave the width of a component implicit in HiFIRRTL programs.
Such unspecified widths are inferred by the Infer Widths pass.
The general rules are,
\begin{itemize}
\item The only allowed width for an explicitly defined width is itself.
\item An implicitly defined width will be set to the largest width that can
  be found through connect statements.
\item The last connection semantics is not considered at this stage,
  all connect statements are taken into account.
\end{itemize}

Following the rules, we first introduce an additional map $wm$, which
maps from variables to types. Then we scan through the statements in $ss$ to
find components with implicit width and create entries for
them in $wm$. For a statement that declares a component $v$,
the {\tt infer\_width} function will update $wm$ by
  \begin{gather}
    \inferii[]
    {\seqq{ce \vdash {\tt infer\_width} (declare\;v, wm) \rightsquigarrow wm'}}
    {\seqq{ce \vdash {\tt width}(t_v) = 0}} {\seqq{wm':= wm \uplus \{v : t_v\}}}.
    \end{gather}
The width is inferred according to connect statements.
Since the last connect semantics is not considered, each
connection to the element provides a potential width. A function
named {\tt max\_width} compares the potential widths and outputs
the maximum one. For a statement which connects an expression $e$
to a component $v$, the {\tt infer\_width} function will update $wm$ by
  \begin{gather}
    \inferiii[]
    {\seqq{ce \vdash {\tt infer\_width }(v<=e, wm) \rightsquigarrow wm'}}
    {\seqq{wm \vdash v : t_v}}
    {\seqq{ce \vdash t_e:= {\tt type\_of\_expr }(e)}}
    {\seqq{wm':= wm \uplus {v : {\tt max\_width }(t_v,t_e)}}}.
    \end{gather}
After applying {\tt infer\_width} to all the statements in $ss$,
$wm$ contains the inferred width for every implicitly defined
component. By merging the information from $wm$ for every entry into $ce$,
the component environment $ce$ contains the
result of the inferred width.

Instead of applying {\tt infer\_width} directly on $ce$, the map
$wm$ contains only type information of the implicit ones that can be
manipulated more efficiently.

%% \begin{description}
%% \item[The input] of InferWidth takes the statments and the component
%%   environment $ce$ that passed from the InferType.  In $ce$ the
%%   unspecified widths are recorded by zero width .  The statements
%%   are assumed to contain only legal connections, which means the two
%%   sides of the connection have equivalent type (mentioned?) and the
%%   flow is from a smaller width to a lager one.
%% \item[The output] of InferWidth give a new component environment $ce'$
%%   where all unspecified types have been infered.  Still it is possible
%%   that after this pass, signals are infered to zero width.  The overall
%%   compilation will use other passes to deal with zero width, which is
%%   not concerned here.
%% \end{description}

%% To state the correctness of InferWidth that implement in Coq, we first
%% recall the semantics of width infering.  There are statements that
%% declare new component, or the statements that define connections. For
%% those components declared with explicit widths, we state that the
%% widths will be kept unchange after the pass.  The implicit width
%% declarations will be collected and await to be assigned to a proper width.
%% The connections give hints to the infering process. For example,
%% \begin{lstlisting}
%%   wire c: SInt
%%   c <= shl(SInt("b-100101100101"), 1)
%% \end{lstlisting}
%% the width of wire $c$ is implicit defined. The connection from an
%% shift-left expression to $c$ will provide the protential width for $c$.
%% According to the rules, the width of c now is 13, because the
%% shift-left expression is of type \lstinline!SInt<13>!. Before applying last
%% connect semantics, the module may contain more than one connections to
%% one target. For example,
%% \begin{lstlisting}
%%   wire c: SInt
%%   wire d: SInt<14>
%%   c <= shl(SInt("b-100101100101"), 1)
%%   c <= d
%% \end{lstlisting}
%% Now, wire $c$ has the width as same as $d$, which is 14. The latter connection
%% from d gives a larger width to the destination $c$. 

\subsubsection{Formalization of InferResets}
In HiFIRRTL, the type \prettyll{Reset} can be used to indicate an
abstract reset signal. The Infer Reset pass is responsible for
inferring the concrete type (asynchronous or synchronous) of the abstract one based on the
connection context. The general rules are,
\begin{itemize}
\item Abstract resets driven by only asynchronous resets will be inferred as asynchronous.
\item Abstract resets that have been invalidated or driven by synchronous signals will be inferred as synchronous.
\item Abstract resets driven by both synchronous and asynchronous resets will be an error.
\end{itemize}

Similar to the Infer Widths pass, a temporary map $rm$ is introduced to
handle abstract resets exclusively. It maps from components to reset
types. The map $rm$ is first initialized with all components of type \prettyll{Reset}
by \eqref{rule-infer-reset-init}.
%% scanning through the statements in $ss$
%% for abstract reset types to be added to $rm$.
%% \david{I am not sure I understand the previous sentence:
%% is $rm$ filled based on $ss$ or based on $ce$?
%% Should it perhaps be:
%% $rm$ is first initialized with all (sub)components of type \prettyll{Reset} by rule \eqref{rule-infer-reset-init},
%% and then the pass scans through the statements in $ss$
%% to decide whether each one of them should become asynchronous or synchronous by rule \eqref{rule-infer-reset-pass}.}
  \begin{gather}
  \label{rule-infer-reset-init}
    \inferii[]
    {\seqq{ce \vdash {\tt infer\_reset} (declare\; v, rm) \rightsquigarrow rm'}}
    {\seqq{ce \vdash t_v : \mbox{Reset}}} {\seqq{rm':= rm \uplus \{v : t_v\}}}.
    \end{gather}
It then continues on the connect
statements to find the inference clues,
checking that the type of the driving signal and the type found in the $rm$ do not conflict.
If they are the same concrete type or at least one of them is abstract,
the inferring continues without error. 
  \begin{gather}
    \label{rule-infer-reset-pass}
    \inferiv[]
    {\seqq{ce \vdash {\tt infer\_reset} (v<=e, rm) \rightsquigarrow rm'}}
    {\seqq{rm \vdash v:t_v}} {\seqq{ce\vdash t_e:={\tt type\_of\_expr}(e)}}
    {\seqq{{\tt \neg conflict} (t_v,t_e)}}
    {\seqq{rm':= rm \uplus \{v : t_e\}}}.
    \end{gather}
Otherwise, the {\tt infer\_reset} reports an error in the map $rm$ and stops.
  \begin{gather}
    \inferiv[]
    {\seqq{ce \vdash {\tt infer\_reset} (v<=e, rm) \rightsquigarrow rm'}}
    {\seqq{rm \vdash v:t_v}} {\seqq{ce\vdash t_e:={\tt type\_of\_expr}(e)}}
    {\seqq{{\tt conflict} (t_v,t_e)}}
    {\seqq{rm':= \bot}}.
    \end{gather}
Here, we consider the case of \prettyll{is invalid} statements as same as a connect statement from a synchronous signal.
If there is no error reported, the inferred reset types stored in the temporary map $rm$ can be
merged into the component environment $ce$ using the {\tt map2} function.
The resulting $ce$ contains the inferred concrete reset types.

\subsubsection{Formalization of ExpandConnects}
In HiFIRRTL, connections can be made between aggregate types.  These
connections can be either full or partial.  The purpose of this pass
is to break down connections between aggregate types and produce
individual ones of corresponding ground types. The general rules are,
\begin{itemize}
\item Connections between ground types can not be expanded.
\item Connections between aggregate types will be expanded recursively.
\end{itemize}

The pass focuses on connect statements in $ss$.
The left-hand side
of a connection is a reference to a component, while the right-hand side
is an expression. In order to obtain expanded
connections, the expression needs to be recursively expanded.
Only expressions of multiplexers \prettyll{mux},
\prettyll{validif} and
references \prettyll{r} can be of aggregate types. The other
expressions, such as binary operations, only operate and result in
ground types. Therefore they do not require expanding.
Expanding a \prettyll{mux} expression is done by,
  \begin{gather}
  \label{expand-mux}
    \infer[]
    {\seqq{ce\vdash{\tt expand\_expr} (\mbox{mux}(e_c,e_t,e_f)) \rightsquigarrow es_{mux}}}
    {\seqq{es_{mux}:=[\mbox{mux}(e_c,t,f)\,|\,t \leftarrow {\tt expand\_expr}(e_t),\,f \leftarrow {\tt expand\_expr}(e_f)]}}.
    \end{gather}
It takes the conditional expression $e_c$ and the true/false branch
expressions $e_t$ and $e_f$ from the original \prettyll{mux} expression, and then produces a list $es_{mux}$
of \prettyll{mux} that combines the lists of expanded $e_t$ and $e_f$
with the same condition $e_c$.
Similarly, expanding a \prettyll{validif} expression produces a list of \prettyll{validif} expressions in $es_{validif}$,
  \begin{gather}
  \label{expand-validif}
    \infer[]
    {\seqq{ce\vdash{\tt expand\_expr} (\mbox{validif}(e_c,e)) \rightsquigarrow es_{validif}}}
    {\seqq{es_{validif}:=[\mbox{validif}(e_c,t)\,|\,t \leftarrow {\tt expand\_expr}(e)]}}.
    \end{gather}
Expanding a references is relatively straightforward.
It create a list of references to the sub-elements according to the
type of the component.
After producing the list of references on the left-hand side and the
list of expressions on the right-hand side, new connect statements can
be generated pairwise from the two by
  \begin{gather}
  \label{expand-cnct}
    \infer[]
    {\seqq{ce\vdash{\tt expand\_cnct} (v <= e) \rightsquigarrow ss_{cnct}}}
    {\seqq{ ss_{cnct}:=[lhs<=rhs\,|\,lhs \leftarrow {\tt expand\_ref}(v),\,rhs \leftarrow {\tt expand\_expr}(e)]}}.
    \end{gather}

The \prettyll{is invalid} statement is treat as the same as the connect statement,
which connects a component with an ``unknown'' signal at this point.
After processing an \prettyll{is invalid} statement for a component,
we obtain a list of \prettyll{is invalid} statements for its sub-elements.



%%%%%%%% remove resets removed
%%%%%%%% remove resets removed
\hide{
\subsubsection{``Remove Resets''}
The Remove Resets pass runs after the other passes, and removes
redundant reset that are synchronous. This is a optimization step that
can potentially improve the performance by removing unnecessary ones.
A synchronous reset is considered redundant if it is only used to
initialize a register.
}
%%%%%%%% remove resets removed
%%%%%%%% remove resets removed
%% For example,
%% \begin{lstlisting}
%% module MyModule :
%%   input myinput: {a: UInt<1> [1]}
%%   output myoutput: {a: UInt<1> [2], b: UInt<1>}
%%   myoutput <- myinput
%% \end{lstlisting}
%% the connection between myinput and myoutput is a partial connection of
%% aggregate types. It can then be expand through ExpandConnects pass and get,
%% \begin{lstlisting}
%% module MyModule :
%%   input myinput_a: UInt<1>
%%   output myoutput_a_0: UInt<1>
%%   output myoutput_a_1: UInt<1>
%%   output myoutput_b: UInt<1>
%%   myoutput_a_0 <= myinput_a_0
%% \end{lstlisting}

%% \begin{description}
%% \item[The input] of ExpandConnects consists of the sequence of
%%   statments representing the original program, the result component
%%   enviroment $ce3$ from InferWidths pass. The $ce3$ collects the
%%   component identifies to their types.
%% \item[The output] of ExpandConnects contains a program in which all
%%   connections between aggregate types are replaced by connections of
%%   their sub elements and a corresponding $ce4$. In $ce4$, new
%%   identifiers added and their types are consistent to the flattened
%%   aggregate type. As the aboved example, ouput port \lstinline!myport!
%%   is replaced by 3 output ports of ground types.
%% \end{description}

\subsubsection{Formalization of ExpandWhens}
This is the main pass that ensures that every component in a module is connected to at most once:
it replaces connect statements within a branch of a \prettyll{when} statement by a single connect statement
and it removes non-last (redundant) connect statements.
This is done by scanning through the statement sequence of the module:
declaration statements are copied without change,
and connect and \prettyll{is invalid} statements update a connect map
(so redundant connects are overwritten in the map).
\begin{gather}
    \label{rule-expandwhen-declare}
    \infer[]
    {\seqq{{\tt expand\_branch}(ss ; \mathit{declare}~v, cm_0) = (ss_\mathrm{dcl} ; \mathit{declare}~v, cm[{\tt not~connected}/v])}}
    {\seqq{{\tt expand\_branch}(ss, cm_0) = (ss_\mathrm{dcl}, cm)}}
\\[1ex]
    \label{rule-expandwhen-connect}
    \infer[]
    {\seqq{{\tt expand\_branch}(ss ; v <= e, cm_0) = (ss_\mathrm{dcl}, cm [e/v])}}
    {\seqq{{\tt expand\_branch}(ss, cm_0) = (ss_\mathrm{dcl}, cm)}}
\\[1ex]
    \label{rule-expandwhen-invalid}
    \infer[]
    {\seqq{{\tt expand\_branch}(ss ; v \text{ \prettyll{is invalid}}, cm_0) = (ss_\mathrm{dcl}, cm [{\tt invalid}/v])}}
    {\seqq{{\tt expand\_branch}(ss, cm_0) = (ss_\mathrm{dcl}, cm)}}
\end{gather}
In contrast to the operational LoFIRRTL semantics defined in Subsection~\ref{sec-lofirrtl-sem},
the connect statements do not need to be ordered topologically.
Therefore, we cannot evaluate expression $e$ in rule~\ref{rule-expandwhen-connect} immediately
but copy it as a syntactical object into the connect map.
This also implies that we cannot use the (register) state of a configuration in these rules.

For \prettyll{when} statements,
the true and false branches are similarly split
and the two resulting connect maps are then combined into one,
using multiplexer or \prettyll{validif} expressions where necessary.
\begin{gather}
    \label{rule-expandwhen-when}
    \inferii[]
    {\seqq{{\tt expand\_branch}(ss ; \text{\prettyll{when c then} } ss^\mathsf{T} \text{ \prettyll{else} } ss^\mathsf{F}) =
        (ss_\mathrm{dcl} ; ss_\mathrm{dcl}^\mathsf{T} ; ss_\mathrm{dcl}^\mathsf{F}, {\tt combine\_branches}(cm^\mathsf{T}, cm^\mathsf{F}))}}
    {\seqq{{\tt expand\_branch}(ss, cm_0) = (ss_\mathrm{dcl}, cm)}}
    {\begin{gathered}[b] \seqq{{\tt expand\_branch}(ss^\mathsf{T}, cm) = (ss_\mathrm{dcl}^\mathsf{T}, cm^\mathsf{T})} \\
                      \seqq{{\tt expand\_branch}(ss^\mathsf{F}, cm) = (ss_\mathrm{dcl}^\mathsf{F}, cm^\mathsf{F})} \end{gathered}}
\end{gather}
The function {\tt combine\_branches} constructs a connect map by the following rules.
If both branches assign the same value, no multiplexer is needed:
{\small\begin{gather*}
    \infer[]
    {{\tt combine\_branches}(cm^\mathsf{T},cm^\mathsf{F})(v) = e}
    {cm^\mathsf{T}(v) = cm^\mathsf{F}(v) = e}
\qquad
    \infer[]
    {{\tt combine\_branches}(cm^\mathsf{T},cm^\mathsf{F})(v) = {\tt invalid}}
    {cm^\mathsf{T}(v) = cm^\mathsf{F}(v) = {\tt invalid}}
\end{gather*}}
If the two are different, insert a multiplexer of a \prettyll{validif} expression:
{\small\begin{gather*}
    \inferiii[]
    {{\tt combine\_branches}(cm^\mathsf{T},cm^\mathsf{F})(v) = \text{\prettyll{mux}}(\text{\prettyll{c}},e^\mathsf{T},e^\mathsf{F})}
    {cm^\mathsf{T}(v) = e^\mathsf{T}}
    {cm^\mathsf{F}(v) = e^\mathsf{F}}
    {e^\mathsf{T} \not= e^\mathsf{F}}
\qquad
    \inferii[]
    {{\tt combine\_branches}(cm^\mathsf{T},cm^\mathsf{F})(v) = \text{\prettyll{validif}}(c,e^\mathsf{T})}
    {cm^\mathsf{T}(v) = e^\mathsf{T}}
    {cm^\mathsf{F}(v) = {\tt invalid}}
\end{gather*}}
If only one branch declared the component, ignore the condition;
if at least one branch did declare the component but erroneously did not connect it,
register that it is still not (always) connected:
{\small\begin{gather*}
    \inferii[]
    {{\tt combine\_branches}(cm^\mathsf{T},cm^\mathsf{F})(v) = e^\mathsf{T}}
    {cm^\mathsf{T}(v) = e^\mathsf{T}}
    {cm^\mathsf{F}(v) \text{ is undefined}}
\qquad
    \infer[]
    {{\tt combine\_branches}(cm^\mathsf{T},cm^\mathsf{F})(v) = {\tt not~connected}}
    {cm^\mathsf{F}(v) = {\tt not~connected}}
\end{gather*}}
For the last three rules, a symmetric variant exists where the role of the true and false branches are swapped.

Finally, to initialize, the empty statement sequence does not change a connect map:
{\small\begin{gather*}
    \infer{\seqq{{\tt expand\_branch}(\emptyset, cm_0) = (\emptyset, cm_0)}}
    {} % no premise
\end{gather*}}
The connect map generated by {\tt expand\_branch} can easily changed back into a list of connect and \prettyll{is invalid} statements
(possibly after topologically sorting them),
similar to the example in Subsection~\ref{subsec-lowering-transformations}.
During this change back, if some component is assigned value {\tt not connected},
an appropriate error message can be generated.

\subsubsection{Formalization of LowerTypes}
The Lower Types pass is primarily aimed at transform aggregate types
into ground types.  The pass replaces certain defined components with
aggregate types by a new set of components defined with ground types.
The general rules are,
\begin{itemize}
\item Components defined with ground types do not need to be modified.
\item Except for memories and module instances, components defined with aggregate types need to be replaced by
      a group of coordinate components that have ground types.
\end{itemize}
The paired identifier is introduced $(id,id)$, in which the
first field represents the base reference adapted from the original
component identifier and the second field indicates the
sub-element offset computed from the aggregate type.
The original components defined with aggregate types are replaced by a new
set of components defined with paired identifiers and of ground types.
The process can be formalized as follow:
  {\small\begin{gather}
    \inferiv[]
    {\seqq{ce\vdash{\tt lower\_types}(declare\;v) \rightsquigarrow (ss',ce')}}
    {\seqq{ce\vdash v: tc_v}}
    {\seqq{{\tt is\_aggr}(tc_v)}}
    {\seqq{ ss' :={\tt flatten}(v,tc_v)}}
    {\seqq{ce':=ce\uplus \{(v,o_1): tc_{v1},\dots,(v,o_n):tc_{vn}\}}}.
    \end{gather}}
The {\tt flatten} function generates new statements that define
components in ground types with new identifiers, such as
$(v,o_1)$. The type and component information contained in $tc_v$
are used to determine the number of the certain new components.
After applying the Expand Connections pass,
connect statements are already in ground types. The references in
these statements need to be replaced by the corresponding paired
identifiers. As a result, the statements $ss$ in the circuit
diagram are replaced by the new $ss'$,
and the components environment $ce$ are updated to reflect the new ground types
and paired identifiers.

