%!TEX root = main.tex

Compilers are among the most fundamental and widely-used
software, based on which programs in high-level programming languages
are translated into low-level code. Well-known open source compilers include GCC~\cite{GCC} for C and C++ programs and LLVM~\cite{LLVM}.
Compilers for main-stream programming languages like C usually include many compilation passes and optimizations. As a result of their inherent complexity,
there remain massive bugs in compilers, despite intensive testing efforts~\cite{LeAS14,ZhangSS17}.
Even worse, compiler bugs could silently entail incorrect executions in low-level programs and lead to catastrophic consequences
for safety-critical applications, and it is extremely difficult to determine whether
an execution failure is caused by defects in the programs or
bugs in compilers. This motivates researchers to apply formal verification techniques to guarantee the correctness of compilers.

Formally verified compilers have been built for several realistic programming languages, such as CompCert for C~\cite{Leroy09,kastner2018compcert,Leroy2006,Leroy-BKSPF-2016,BlazyDL06,Kastner-LBSSF-2017},
V\'{e}lus for Lustre~\cite{BourkeBDLPR17}, Vellvm~\cite{ZhaoNMZ12,ZhaoNMZ13} and Crellvm~\cite{KangKSLPSKCCHY18} for the LLVM Intermediate Representation,
and CakeML for ML~\cite{KumarMNO14,LoowKTMNAF19,TanMKFON19,OwensNKMT17}.
Such compilers come with a mathematical, machine-checked proof
that compilation passes preserve program semantics,
hence strengthening the correctness guarantees that can be obtained by formal verification of high-level programs.

The success of formally verified compilers on software programming languages has been ported to hardware description languages (HDLs)
very recently, e.g., the extension of CakeML (called Lutsig) for compiling %register-transfer language Verilog to gate-level netlist~\cite{Loow21,loowphd21}
a subset of register-transfer level (RTL) Verilog programs to netlist level programs for field-programmable gate array (FPGA)\cite{Loow21,loowphd21},
and the extension of CompCert (called Vericert) for compiling %compiling C programs to Verilog programs~\cite{HerklotzPRW21}.
a subset of C programs to RTL Verilog programs~\cite{HerklotzPRW21}.
Following this promising research direction, in this work, we establish for the first time a framework that formally reasons about programs in FIRRTL~\cite{firrtl-spec} (Flexible Intermediate Representation for RTL) and its lowering transformations (aka compilation passes).

FIRRTL originated from the high-level hardware construction language Chisel~\cite{BachrachVRLWAWA12}.
Chisel is a domain specific language embedded in Scala that comprises a library of special class definitions, predefined objects, and usage conventions for writing highly-parameterized circuit design generators.
High-level hardware generators in Chisel are developed by manipulating circuit components using Scala functions,
and their interfaces are encoded in Scala types.
Circuits can also be constructed by leveraging Scala's functional and object-oriented programming features, enabling parameterized, modular, and reusable hardware generators that improve hardware design productivity and robustness.
Chisel has been successfully used to build RISC-V processes such as Rocket Chip~\cite{RocketChip}, RISC-V BOOM \cite{Boom}, NutShell~\cite{NutShell},
XiangShan~\cite{xiangshan,Xiangshan-github}, and so on.
%SiFive Core IP portfolio~\cite{SiFive},
%and Google Edge TPU~\cite{EdgeTPU}.

The FIRRTL intermediate representation (IR) was proposed for the compilation of Chisel programs~\cite{IzraelevitzKLLW17}, analogous to the role of LLVM IR~\cite{LLVM} for compiling C programs.
FIRRTL features first-class support for high-level constructs such as vector types, bundle types, conditional statements, connects, partial connects,
and modules so that Chisel programs can easily be translated into FIRRTL.
%
High-level synthesis is performed on FIRRTL by a sequence of compilation passes  (aka lowering transformations) and optimizations,
resulting in a most restricted form that is close to RTL Verilog, called low FIRRTL (abbreviated as LoFIRRTL).
For clarity, let us call FIRRTL with high-level constructs \emph{high FIRRTL} (abbreviated as HiFIRRTL).
While FIRRTL was designed to compile Chisel programs, it is not tied to Chisel. Other hardware description languages can also be compiled into FIRRTL and reuse the majority of the FIRRTL compilation passes. For instance, FIRRTL has been integrated into the CIRCT project~\cite{CIRCT} as one of its core dialects.


%%%%%% resolved conflicts
\hide{
The FIRRTL intermediate representation was proposed for the compilation of Chisel programs~\cite{IzraelevitzKLLW17},
analogous to the role of the LLVM IR~\cite{LLVM} for compiling C programs.
FIRRTL features first-class support for high-level constructs
such as vector types, bundle types, conditional statements, connects, partial connects,
and modules so that Chisel programs can easily be translated into FIRRTL.
}
%

The FIRRTL compiler suffers from lots of bugs due to its inherent complexity. For instance, there are 655 issues in its github repository, among which
99 have been labeled as bugs~\cite{FIRRTLissues}. This motivates us to investigate the formalization and verification of FIRRTL and its lowering transformations.
%
%For instance, the width of \lstinline!dshl(e1,e2)! inferred by the ``Infer Width'' compilation pass
%is less than the width of \lstinline!e1!~\footnote{\url{https://github.com/chipsalliance/firrtl/issues/2296}}, where \lstinline!dshl! is a dynamic shift left operation
%which shifts the bits in \lstinline!e1! towards the
%most significant bit \lstinline!e2! places and \lstinline!e2! zeroes are shifted in to the least significant bits.
%The ``ExpandWhens'' compilation pass may introduce unexpected \lstinline!VOID!, leading to a match error in
%the ``InferTypes'' compilation pass~\footnote{\url{https://github.com/chipsalliance/firrtl/issues/852}}.
%
Nevertheless, partially due to the fact that Chisel and FIRRTL are relatively new and the ecosystem around them is still evolving, formalization and verification of FIRRTL IR and its lowering transformations are largely missing, with an exception \cite{HarrisonHA21}. In~\cite{HarrisonHA21}, a coinductive denotational semantics was proposed for \fu{a subset of} LoFIRRTL, based on a formalism called device calculus. Nevertheless, while Harrison et al. made progress in the formalization of LoFIRRTL in~\cite{HarrisonHA21}, a framework where  programs in HiFIRRTL and its lowering transformations can be formally reasoned about is still elusive. The establishment of such a framework would require
%\begin{itemize}
%\item
a mechanized operational semantics of LoFIRRTL,
%
%\item
validation of the conformance of the semantics to the FIRRTL reference interpreter,
%
%\item
formalization of HiFIRRTL and its lowering transformations,
%
%\item
as well as specification and verification of the correctness of its lowering transformations.
%\end{itemize}

In this work, we fill in this gap and provide such a framework. Specifically, we make the following contributions.
\begin{itemize}
\item Define an executable operational semantics for LoFIRRTL in Coq. The operational semantics is based on a concept of pseudo-single-static-assignment form of LoFIRRTL programs. Our formalization covers the generic LoFIRRTL for digit circuit, with the exception of the memory semantics, which is ambiguous under active debate. To avoid any potential inconsistencies, we choose to omit it from our representation and focus on the others that have well-defined semantics. We introduce the concept of configurations to trace the execution of the operational semantics and to provide evidences for type safe executions.
%
\item Validate the operational semantics against Treadle, the main-stream FIRRTL reference interpreter, on a benchmark suite comprising 120 LoFIRRTL programs.
%
\item Introduce a concept of circuit diagrams  and formalize the syntax of HiFIRRTL and its six lowering transformations in Coq, namely, ``InferTypes'', ``InferWidths'', ``InferResets'', ``ExpandConnects'', ``ExpandWhens'', ``LowerTypes''.
Those selected transformations are believed to be the most involved ones. The formalization in Coq follows carefully the restrictions provided by the official specification. The use of circuit diagrams in the formalization helps to keep track of the facts before and after transformations, which is useful for the verifying the functional correctness.
%
%% \item Specify and verify the functional correctness of the two most involved lowering transformations \fu{in Coq}, namely ``ExpandConnects'' and ``ExpandWhens''.
\item Specify and verify the functional correctness of the two most intricate and impactful lowering transformations in Coq, namely ``ExpandConnects'' and ``ExpandWhens''. They involve significant changes to the original HiFIRRTL program, such as creating new connections and thus potentially impact on the functionality of the circuit. The verification ensures the functional correctness that these transformations preserve the behaviors of the original programs by taking the type equivalence and the (conditional) last connect semantics into account.
\end{itemize}

The rest of this paper is organized as follows. Section~\ref{sec-firrtl} gives an informal introduction to FIRRTL and its lowering transformations. Section~\ref{sec-form-low-firrtl} is devoted to the  operational semantics of LoFIRRTL as well as its validation against Treadle. Section~\ref{sec-hi-firrtl-transform} formalizes HiFIRRTL and its lowering transformations. Section~\ref{sec-spec-verif-expand} specifies and verifies ``ExpandConnects'' and ``ExpandWhens''. We discuss related work in Section~\ref{sec-related-work}
and conclude this paper in Section~\ref{sec-conc}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\hide{
Surprisingly, a formal operational semantics of FIRRTL is still missing and
the FIRRTL compiler suffers from lots of bugs due to the complexity of FIRRTL transformations.
In fact,  there are 649 issues in its github repository, among which
99 have been labeled as bugs~\cite{FIRRTLissures}.
For instance, the width of \lstinline!dshl(e1,e2)! inferred by the ``Infer Width'' compilation pass
is less than the width of \lstinline!e1!,\footnote{\url{https://github.com/chipsalliance/firrtl/issues/2296}} where \lstinline!dshl! is a dynamic shift left operation
which shifts the bits in \lstinline!e1! towards the
most significant bit \lstinline!e2! places and \lstinline!e2! zeroes are shifted in to the least significant bits.
The ``ExpandWhens'' compilation pass may introduce unexpected \lstinline!VOID!, leading to a match error in
the ``InferTypes'' compilation pass\footnote{\url{https://github.com/chipsalliance/firrtl/issues/852}}.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%We define for the first time the executable operational semantics of low FIRRTL.
%present the first formal operational semantics of the hardware description language FIRRTL (Flexible Intermediate Representation for RTL)~\cite{firrtl-spec} and verify two key compilation passes of FIRRTL programs~\cite{firrtlcompiler,IzraelevitzKLLW17}, namely ``Infer Width'' and ``Expand Whens'',
%which infers bit widths of signals
%and replaces conditional statements with multiplexers, respectively.

%\fu{In this work, we make the first attempt to fill the above gap. We tbd. Describe what we have done, e.g., challenges of formalizing
%operational semantics of FIRRTL and transformations, how we address these challenges,
%results of our work.}
