%!TEX root = main.tex

%The proof of the functional correctness of ExpandConnects and ExpandWhens.

In this section, 
we first present the specifications of the ExpandConnects and ExpandWhens passes, 
then prove their correctness, namely, the Coq implementations conform to their specifications.


The correctness proof for the ExpandWhens pass compares the Coq implementation with (a formalization of) the output specified above.
It requires a mutual induction over statements and statement sequences
because the \lstinline~when~ statement contains two statement sequences,
but the proof is more long than complicated. \fu{May move into the next subsection}

\subsection{Correctness of ExpandConnects}
Recall the the ExpandConnects pass is aimed to decompose the connect statements of aggregate
types to ground type connections.
To prove the functional correctness of the ExpandConnects pass, we need to ensure
that the sub elements of the components are connected to their equivalent ground types.
Let $ce^{in}$ be the input component environment,
%% such that $ce^{in}$ is the result of the InferTypes pass,
and $ss^{in}$ be the list of connect statements in the input circuit
diagram $D_M^{in}$.
%% Let $ss^{out}$ be the output list of connect
%% statements after applying the Expand Connections pass.
The input
$D_M^{in}$ is required to be the result of the InferTypes pass, in order
to make sure that the connections are made between equivalent types.

We specify the correctness based on the types of the left-hand side and the right-hand side
expressions in the resulting connect statements.
First, we state that the type equivalence holds for the output statements generated by the ExpandConnects pass
as,
\begin{lemma}\label{expand-conct-type1-ss}
$\forall ce^{in}\; ss^{in},\, ce^{in} \models TypeEquivs\, ss^{in} \Rightarrow TypeEquivs\, [{\tt expand\_cnct}\,s\,|\,s\leftarrow ss^{in}]).$
\end{lemma}
Intuitively, it states that if the type equivalence predicate holds for all the
connect statements in $ss^{in}$,
it holds for the statements produced by applying the {\tt expand\_cnct} \eqref{expand-cnct} function on all the inputs.
More specifically, the type equivalence predicate $TypeEquivs$ holds for a list of statements if
for each individual comparison of the types on the two sides of a connection, the type equivalence
predicate $TypeEquiv$ holds.
If there are $n$ statements $s_1,s_2,\dots ,s_n$ in the sequence $ss$ of statements,
then $TypeEquivs\;ss\,\!=\,\!\bigwedge_{i=0}^{n} TypeEquiv\;s_i$.

The proof can be achieved by induction on the list of statements,
assuming we have the proof for expanding one statement.
Since the function {\tt expand\_cnct} only affect connect statements,
we can state the following lemma:
\begin{lemma}\label{expand-conct-type1-s}
$\forall ce^{in}\;v\;e,\, ce^{in} \models TypeEquiv\,(v<=e) \Rightarrow TypeEquivs\, ({\tt expand\_cnct}\,(v<=e)).$
\end{lemma}
If the predicate $TypeEquiv$ holds for a connect statement $v<=e$ under a component environment,
then the types of $v$ and $e$ are equivalent.
Lemma \ref{expand-conct-type1-s} can be further simplified as,
\begin{lemma}\label{expand-conct-type1-s}
$\forall ce^{in}\;v\;e,\, ce^{in} \models Equiv\,({\tt typeof}\,v,\;{\tt typeof}\,e) \Rightarrow$
$Equivs\,[({\tt typeof}\,l,\,{\tt typeof}\,r)\,|\,l \leftarrow{\tt expand\_expr}\,v,\;r\leftarrow {\tt expand\_expr}\,e]$.
\end{lemma}
If there are $n$ pairs of types $(t_1^l,t_1^r),(t_2^l,t_2^r),\dots ,(t_n^l,t_n^r)$ in the list $ts$,
then $Equivs$ holds for $ts$ if
$\bigwedge_{i=0}^{n} Equiv\;(t_i^l\,t_i^r)$.
In Lemma \ref{expand-conct-type1-s}, the list of pairs are the types obtained by
expanding the reference on the left-hand side
and expanding the expression on the right-hand side, respectively.
If we can build the relation between the aggregate type of an expression and the types of
the corresponding expanded expressions, the above lemma can be proved.
Here we introduce a function {\tt dec\_type} that can decompose the aggregate types
by its structure to get a list of ground types.
A related lemma is,
\begin{lemma}\label{dec-type}
$\forall ce^{in}\, t_1\, t_2,\,ce^{in}\models Equiv\,(t_1,\;t_2) \Rightarrow$
$Equivs\,[(x,\,y)\,|\,x \leftarrow{\tt dec\_type}\,t_1,\;y\leftarrow {\tt dec\_type}\,$ $t_2]$
\end{lemma}
It states that, decomposing a pair of equivalent aggregate types produces two pairwise equivalent type lists.
If we use the {\tt dec\_type} function on the aggregate type of the expression  to be expanded, it
returns the ground types that are identical to the ones produced by {\tt expand\_expr}, as in the following lemma.
\begin{lemma}\label{expand-expr-dec-type}
$\forall ce^{in}\, e,\,ce^{in}\models[{\tt typeof}\,x\,|\,x\leftarrow{\tt expand\_expr}\,e] = {\tt dec\_type}\,({\tt typeof}\,e)$
\end{lemma}
By rewriting Lemma \ref{expand-expr-dec-type} in the goal of Lemma \ref{expand-conct-type1-s}, the proof goal is change to
$Equivs\,[(x,\,y)\,|\,x\leftarrow{\tt dec\_type}\;({\tt typeof}\,v),\;y\leftarrow {\tt dec\_type}\,({\tt typeof}\,e)]$.
Now we can apply Lemma \ref{dec-type} with assigning $t_1$ and $t_2$ with ${\tt typeof}\,v$ and ${\tt typeof}\,e$
to complete the proof.

Secondly, we state that all the expressions after applying the ExpandConnects pass
have ground types.
\begin{lemma}
$\forall ce ^{in}\,ss^{in},\,ce^{in}\models AllGround ({\tt expand\_cnct}\,ss^{in})$
\end{lemma}

The above lemma states the predicate $AllGround$ holds for the generated
statements of {\tt expand\_cnct}.
Similarly, the proof goal will be reduced to prove that the $AllGround$ predicate holds for the expanded expressions.
It is rather easy to use Lemma \ref{expand-expr-dec-type} we had for {\tt dec\_type}.
\fu{To relate the types of the expanded expressions are identical to decompose the type of the original expression.}
By induction on the aggregate type constructors, it is able to prove that decomposed types by
the {\tt dec\_type} function will always return ground types.

\xiaomu{Stopped}

\input{verif-firrtl-transform-expandwhens}


%%%%% the formal semantics of HiFIRRTL removed %%%%%%%%%%
%%%%% the formal semantics of HiFIRRTL removed %%%%%%%%%%
\hide{
The formal semantics of HiFIRRTL will be defined by the axiomatic semantics of the compilation passes, plus the operational semantics of LoFIRRTL.
%InferTypes, InferWidths, InferResets, ExpandConnects, ExpandWhens, LowerTypes, and RemoveReset.
%
More precisely, the axiomatic semantics of each compilation pass $\pi$ will be specified by a formula $\varphi_{\pi}(\cdots)$, and
%Note that the sequence of statements in $ss$ will keep unchanged in the compilation passes.
%
%ce1 = φ_InferType(ss) /\
%ce2 = φ_InferWidth(ss, ce1) /\
%ce3 = φ_InferReset(ss, ce2) /\
%(ss1, ce4) = φ_ExpandConnect(ss, ce3) /\
%(ss2, cm) = φ_ExpandWhen(ss1, ce4) /\
%(ss3, ce5, cm1) = φ_LowerTypes(ss2, ce4, cm) /\
%(ss4, ce6, cm2) = φ_RemoveResets(ss3, ce5, cm1)
the semantics of $M$ is defined by the formula
\[
\begin{array}{l c l}
\varphi_M & \eqdef & \varphi_{ResolveKinds}(ss, ce_1)  \wedge  \varphi_{InferTypes}(ss, ce_1, ce_2)\ \wedge \\
&& \varphi_{InferWidths}(ss, ce_2, ce_3) \wedge  \varphi_{InferResets}(ss, ce_3, ce_4)\ \wedge \\
&& \varphi_{ExpandConnects}(ss, ce_4, ss_1, ce_5) \wedge \varphi_{ExpandWhens}(ss_1, ce_5, ss_2, cm)\ \wedge \\
&& \varphi_{LowerTypes}(ss_2, ce_5, cm, ss_3, ce_6, cm_1) \wedge  \varphi_{RemoveReset}(ss_3, ce_6, cm_1, ss_4, ce_7, cm_2).
\end{array}
\]
Note that each formula $\varphi_{\pi}$ actually specifies the graph of a function corresponding to $\pi$. For instance, the formula $\varphi_{InferWidths}(ss, ce_2, ce_3)$ specifies a function that takes $ss$ and $ce_2$ as inputs and produces the output $ce_3$.  The final output $ss_4$ is actually a sequence of statements already in LoFIRRTL form, whose operational semantics has been defined in Section~\ref{sec-lofirrtl-sem}.

In the sequel, we shall illustrate the definition of the semantics of compilation passes by describing $\varphi_{InferWidths}$ and $\varphi_{ExpandWhens}$. The semantics of the other compilation passes can be found in Appendix~\ref{sec-appen-sem-cpass}.

%Let us fix $ss = (s_1, \cdots, s_n)$ in the sequel.

$\varphi_{InferWidths}(ss, ce, ce') \eqdef \varphi_{InferWidths}(ss, ce, ce')$


$\varphi_{ExpandWhens}$

last connect semantics

four states of components $\{\bot, \sim, -\}$ (undefined, defined, invalid),

For $v \in \varset(M)$, we construct an expression $EWRhsExp_{ss, ce, v}(rhs)$ inductively as follows,
and for $i \in [n]$,
\begin{itemize}
\item If $ss = \varepsilon$, then $EWRhsExp_{ss, ce, v}(rhs) = rhs$,
%
\item If $ss = s \cdot tl$, then
\begin{itemize}
\item case $s \equiv v' \mbox{ is invalid}$: if $v = v'$ and $rhs = \bot$ or $\sim$, then $EWRhsExp_{ss, ce, v}(rhs) = EWRhsExp_{tl, v, -}$, otherwise, $EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, rhs}$,
%
\item case $s \equiv \mbox{wire } v': t$: if $v = v'$ and $rhs = \bot$, then $EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, \sim}$, otherwise,  $EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, rhs}$,
%
\item case $s \equiv \mbox{reg } v': t, c$: if $v = v'$ and $rhs = \bot$, then $EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, \sim}$, otherwise, $EWRhsExp_{ss, v}  = EWRhsExp_{tl, v, rhs}$,
%
\item case $s \equiv v'<= e$: if $v = v'$, then $EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, e}$, otherwise,  $EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, rhs}$,
%
\item case $s \equiv \mbox{when } c: sst \mbox{ else } ssf$: suppose $rhst = EWRhsExp_{sst, v, rhs}$ and $rhsf = EWRhsExp_{ssf, v, rhs}$, then
$EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, mux(c, rhst, rhsf)}$,
%
%\item case $s \equiv v' \mbox{ inst of } v''$: if $v = v'$ and $rhs = \bot$, then $EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, \sim}$, otherwise, $EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, rhs}$,
%
\item for the other statements,  $EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, rhs}$.
%\begin{itemize}
%\item if $rhst = \bot$ or $\sim$, then $EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, rhst}$,
%
%\item otherwise, if $rhsf = \bot$, then $EWRhsExp_{ss, v, rhs} = EWRhsExp_{sst, v, rhs}$,
%
%\item if $rhst \neq \bot$ and $rhsf \neq \bot$, then if $rhst = rhsf$, then $EWRhsExp_{ss, v, rhs}  = rhst$, otherwise, $EWRhsExp_{ss, v, rhs} = EWRhsExp_{tl, v, mux(c, rhst, rhsf)}$,
%
%\item
%\end{itemize}
\end{itemize}
\end{itemize}

Finally, we transform $EWRhsExp_{ss, v, rhs}$ into $EWRhsExp'_{ss, v, rhs}$ as follows:
\begin{itemize}
\item If $EWRhsExp_{ss, v, rhs} = mux(c, e_1, e_2)$, then
\begin{itemize}
\item if both $e_1$ or $e_2$ contains occurrences of $\bot, \sim$, then $EWRhsExp'_{ss, v, rhs} = \bot$,
%
\item if both
\end{itemize}
\end{itemize}

Define a recursive function $expandWhens$
}
%%%%% the formal semantics of HiFIRRTL removed %%%%%%%%%%
%%%%% the formal semantics of HiFIRRTL removed %%%%%%%%%%
