%!TEX root = main.tex

%FIRRTL represents the standardized elaborated circuit that the Chisel HDL produces. 
%FIRRTL represents the circuit immediately after Chisel's elaboration but before any circuit simplification. It is designed to resemble the Chisel HDL after all meta-programming has executed. Thus, a user program that makes little use of meta-programming facilities should look almost identical to the generated FIRRTL.

%For this reason, 
FIRRTL has first-class support for high-level constructs such as vector types, bundle types, conditional statements, partial connects, and modules. These high-level constructs are then gradually removed by a sequence of lowering transformations. During each lowering transformation, the circuit is rewritten into an equivalent circuit using simpler, lower-level
constructs. Eventually the circuit is simplified to its most restricted form, resembling a structured netlist, which allows for easy translation to an output language (e.g. Verilog). This form is given the name lowered FIRRTL (LoFIRRTL) and is a strict subset of the full FIRRTL language. For clarity, the original form is called high FIRRTL (HiFIRRTL). 

The syntax of FIRRTL is given in Figure~\ref{fig-firrtl-synt}.  Note that the syntactic rules therein are in Extended Backus–Naur form, where $\{\cdots\}$ denotes repetition (possibly empty) and $[\cdots]$ denotes option, moreover, the terminals are doubly quoted. 

\input{firrtl-syntax}


\begin{itemize}
\item A circuit in FIRRTL consists of modules, each of which include several ports and statements. 
\item Each port has an identifier and types and is either an input or output. 
\item Each statement defines a wire, register, memory, or node, connect (<=) or partial connect (<-), attach (for analog signals), or conditional (when), and so on.  
\item Expressions are widely used in statements. Each expression is an unsigned/signed integer constant, a reference, a multiplexer, a validif expression, or an expression obtained by applying primitive operations on ground types. 
\item A ground type is ``Clock'', ``Reset'', ``AsyncReset'', ``UInt'' (unsigned integer), ``SInt'' (signed integer), ``Analog'', or ``Fixed'' (fixed point number).  
\item An aggregate type is either declared by describing the (possibly nested) fields, or an array (e.g. UInt[10]).
\end{itemize}
Note that some special terminal symbols $newline$, $indent$, and $dedent$ are used to format FIRRTL programs. 

A FIRRTL circuit is defined to be a valid LoFIRRTL circuit if it obeys the following restrictions:
\begin{itemize}
\item All widths must be explicitly defined.
\item The conditional statement is not used.
\item All components are connected to at most once.
\item All components must be declared with a ground type.
\item The partial connect statement is not used.
\end{itemize}
Intuitively, LoFIRRTL is in a flavor similar to static single-assignment (SSA) form of software programs. Programs in SSA form are those where each variable is assigned exactly once, and each variable is defined before it is used.

Besides LoFIRRTL, there is another lowered form of FIRRTL, MidFIRRTL. Nevertheless, in this paper, we ignore MidFIRRTL, since there are no requirements in FIRRTL related to accepting or producing MidFIRRTL.

