%!TEX root = main.tex
Formalizing the operational semantics of LoFIRRTL in Coq not only provides a formal reference of LoFIRRTL,
but only yields an executable interpreter in OCaml, called LoFIRRTL OCaml Interpreter, via the Coq's extraction feature.
The former would greatly facilitate developers' understanding, freeing them from lengthy, ambiguous and elusive documentations.
The latter enables developers' to conveniently test correctness of FIRRTL circuit designs and test conformance of any other
LoFIRRTL interpreter or simulator via difference testing.
In this subsection, to understand the performance and ensure confidence of the formalized operational semantics,
we conduct difference testing between our LoFIRRTL OCaml Interpreter and
the official FIRRTL circuit simulator, called Treadle, which executes LoFIRRTL circuts
based on the earlier scala-based LoFIRRTL Interpreter.


\smallskip\noindent
{\bf Benchmarks}. We mainly collect LoFIRRTL circuits from Treadle, 
i.e., those 55 LoFIRRTL circuits inside the ``chipsalliance/treadle/samples'' directory.
We consider those LoFIRRTL circuits because Treadle is more likely to execute correctly on them so that
the correctness testing of the formalized operational semantics has high confidence. 
Among 55 LoFIRRTL circuits, we exclude 21 LoFIRRTL circuits because they violate the syntax of
LoFIRRTL (i.e., dependency of external modules or the usage of subfield), have no output ports or 
cannot be handled by Treadle due to cyclical connection of ports.
Thus, we collected 34 LoFIRRTL circuits from Treadle.
In addition, we also collect all the four LoFIRRTL circuits from 
\cite{HarrisonHA21} which have been tested using the scala-based LoFIRRTL Interpreter
and a formal Agda-based Interpreter designed for a subset of LoFIRRTL.
In summary, we use 38 LoFIRRTL circuits all of which can be handled by Treadle.
 
\smallskip\noindent
{\bf Testing method}.
Since the input LoFIRRTL programs to our LoFIRRTL OCaml Interpreter should be in the pseudo-SSA form
while all the collected LoFIRRTL circuits are not necessary in the pseudo-SSA form,
to solve this issue, we implement a preprocessor
in OCaml which transforms each LoFIRRTL program into an equivalent one in the pseudo-SSA form on which
our LoFIRRTL OCaml Interpreter can be handled.
Furthermore, LoFIRRTL programs themselves require inputs to execute.
For this purpose, we also implement a deriver in OCaml which first automatically identifies
all the input ports and their types for each LoFIRRTL program and then randomly generates 100 test inputs for
the LoFIRRTL program. Finally, for each given LoFIRRTL program and each its test input,
we compare the equivalence of output ports produced by  our LoFIRRTL OCaml Interpreter and Treadle.


\smallskip\noindent
{\bf Testing results}.

%
%
%
%%Coq's Extraction feature to translate lower firrtl (clk\_steps\_tail\_rec) into an OCaml program that is the Low FIRRTL OCaml Interpreter
%
%%benchmarks: treadle\_lofir (self-contained XXX samples out of YYY samples from the official hardware circuit simulator \textsf{treadle}, )
%%and Accumulator.lo.fir, Combinational.lo.fir, FibCircuit.lo.fir, ShiftRegister.lo.fir (all the benchmarks from ``A Mechanized Semantic Metalanguage for
%%High Level Synthesis").
%
%% treadle例子一共55个，18有external module或者多个module用到subfield（是不符合lofir语法），一致的37个,3
%%7个里还有2个例子是电路本身没有output端口没法比较，还有1个是有循环连接的，treadle那边会直接报一个error，ocaml这边我会写一个判断报错
%
%%Testing:  Treadle vs. Interpreter
%
%Steps: for benchmark, an OCaml driver identifies the inputs and their types, then
%randomly generate 100 tests of the benchmark.
%
%test results are the same
%
%execution time:
