%!TEX root = main.tex

We conduct conformance testing of our OCaml FIRRTL compiler to validate the
consistency with the official MLIR FIRRTL compiler {\sf firtool}, deployed in the CIRCT project.



\smallskip\noindent
{\bf Benchmarks}. To extensively evaluate our OCaml FIRRTL compiler,
we mainly consider the tests of the MLIR FIRRTL compiler {\sf firtool}\footnote{\url{https://github.com/llvm/circt/tree/main/test/firtool}.} and the examples
of the Chisel Book--Digital Design with Chisel~\cite{chisel:book}.
%
There are 121 tests for {\sf firtool}, among which 77 programs are excluded because they are not self-contained (i.e., dependent upon external modules),
or contain undefined semantics for memories, or have no output ports, or have illegal connections, or have probe and instance statements
which are not considered in this work. Thus, we consider the remaining 44 FIRRTL programs.
%
There are 51 Chisel examples in Scala inside the ``schoeberl/chisel-book/src/main/scala'' directory of the Chisel Book.
These 51 Chisel examples are transformed into 62 FIRRTL programs by applying \textsf{ChiselStage}
to each subclass of the class {\tt Module} in Scala. (Note that some Chisel examples contain multiple such subclasses, thus yield multiple FIRRTL programs.)
%
%For those from Treadle are 55 LoFIRRTL circuits inside the ``chipsalliance/treadle/samples'' directory.
%We consider those LoFIRRTL circuits because Treadle is more likely to execute correctly on them so that
%the correctness testing of the formalized operational semantics has high confidence.
%Among 55 LoFIRRTL circuits, we exclude 18 LoFIRRTL circuits because they violate the syntax of
%LoFIRRTL (i.e., dependency of external modules), undefined semantics for memories, have no output ports or
%cannot be handled by Treadle due to cyclical connection of ports.
%Thus, we collected 37 LoFIRRTL circuits from Treadle, among them 3 circuits have instof statement,  which means a nested module structure.
%Beside, we generated 76 LoFIRRTL circuits from book project of Digital Design with Chisel, which containing 51 Chisel examples inside the ``schoeberl/chisel-book/src/main/scala'' directory. The transformation is executed by Chisel stage for each class extends from class $\tt Module$, as it is legal to define multi-class in one Scala program, it is possible to execute 76 LoFIRRTL circuits from 51 Chisel programs. The generated LoFIRRTL circuits including 66 single module circuits and 10 with instance structure.
Besides the above FIRRTL programs, we also collect five issues reported as bugs of the three lowering transformations in the GitHub repository
of CIRCT, four of which is about the InferWidths pass and the remaining one is about the ExpandConnects pass. 
Moreover, one of them related to the InferWidths pass remains unresolved while the other four have been fixed in the latest {\sf firtool}.
We develop 7 FIRRTL programs to reproduce these issues. 
%
In summary, we use 113 (44+62+7) FIRRTL programs as our evaluation subjects. % all of which can be handled by {\sf firtool}.

\smallskip\noindent
{\bf Testing method}.
To validate the consistency between our OCaml FIRRTL compiler and the official MLIR FIRRTL compiler {\sf firtool},
for each FIRRTL program, we should check the functional equivalence (i.e., sequential equivalence) of two LoFIRRTL programs produced by the two compilers.
However, to the best of our knowledge, there is no functional equivalence checkers for LoFIRRTL programs and
it is non-trivial to implement such a tool. To address this issue, we transform
LoFIRRTL programs into Verilog by {\sf firtool} which in turn are transformed into AIGER format~\cite{aiger} via {\sf Yosys}~\cite{Yosys}.
The equivalence of two programs in AIGER format is checked
by applying the CEC engine of the tool {\sf UC Berkeley-ABC}~\cite{MishchenkoCBE06,BraytonM10}.

%Since we want to compare the equivalence of circuit structure generated by our FIRRTL OCaml compiler and {\sf firtool}, we first produce LoFIRRTL program of MLIR syntax and use {\sf firtool} to transform it verilog. Then we use yosys \footnote{\url{https://github.com/YosysHQ/yosys}.} to transform verilog programs to aiger files and compare the equivalence of our compile results and the official ones by berkeley-abc \footnote{\url{https://github.com/berkeley-abc/abc}.}.



\smallskip\noindent
{\bf Testing results}.
The summary of the compiler conformance testing is reported in Table~\ref{tab:conformancetesting}.
Column (\#LOC) shows the average number of locations for each benchmark suit.
Column ({\sf firtool}) shows the average compilation time $t_1$ of {\sf firtool} in millisecond (ms) for each benchmark suit.
Column (Ours) shows the average compilation time $t_2$ of our FIRRTL OCaml compiler in ms for each benchmark suit.
Column (Speedup) is calculated via $\frac{t_1-t_2}{t_2}$ for each benchmark suit.
Column (Verification time) shows the average time $t_v$ of equivalence checking in millisecond (ms) for each benchmark suit.
Column (Result) shows the result of equivalence checking for each benchmark suit, 
where \ding{51} indicates that the two programs in AIGER format are equivalent and \ding{55}(1) indicates
that  {\sf firtool} fails to compile one FIRRTL program.
The last row (Average) shows the average results over all 113 benchmarks when they are available. 
Detailed results are given in Appendix~\ref{sec-detailed-results} due to space limits.

\begin{table}[t]
\caption{Results of the compiler conformance testing}\small %\footnotesize\vspace*{-3mm}
\label{tab:conformancetesting}
\scalebox{0.85}{\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
\multirow{2}{*}{\bf Benchmark suit} & \multirow{2}{*}{\bf \#LOC} & \multicolumn{2}{c|}{\bf Compilation time (ms)}&\multirow{2}{*}{\bf Speedup}  & \multirow{2}{*}{\bf Verification time} & \multirow{2}{*}{\bf Result} \\ \cline{3-4}
                              &                        & {\sf firtool}  &  {\bf Ours}  &     & & \\ \hline \hline
      CIRCT (44 programs) & 36.43  & 1.10  & 0.32  & 2.50  & 0.00 & \ding{51} \\ \hline
         Chisel Book (62 programs) & 79.13 & 9.89  & 7.68  &  0.29 & 0.00 & \ding{51} \\ \hline
        Issue (7 programs) & 44.29 & 1.35  & 0.044  & 29.68 & 0.00 & \ding{55}(1) \\ \hline \hline
        \textbf{Average} &  \textbf{60.35} & 	\textbf{5.94} & 	\textbf{4.34} & 	\textbf{0.37} & \textbf{0.00} &  \\ \hline
 \end{tabular}}
\end{table}
 	 	 	
 
Overall, our OCaml FIRRTL compiler successfully compiled all the 113 FIRRTL programs, all of which
are equivalent to the programs produced by {\sf firtool} except for one FIRRTL program that reproduces an unsolved issue of {\sf firtool}.
After an in-depth investigation, we found that {\sf firtool} uses a lightweight constraint solving based algorithm to infer bit widths which wrongly detected a loop in the program.
In fact, the program does not have a loop and its bit widths should be successfully inferred. 
Our OCaml FIRRTL compiler successfully compiled this program because we implement a topological sorting algorithm in Coq to infer bit widths 
so that the FIRRTL program can be successfully lowered.

 
 %all the 106 benchmarks from CIRCT and Chiselbook and produces the same circuit structure as {\sf firtool}. In the official CIRCT database, there are five bugs associated with these three passes, four of which is about inferwidths and one is about expandconnects. Four of these have been resolved, ensuring they can be executed correctly in the latest {\sf firtool}. One of the bugs related to inferwidths remains unresolved. This issue arises from the fact that {\sf firtool} uses constraint solver to infer bit widths, and their algorithm detected a loop in the circuit. However, the circuit does not form a loop and should be solvable. Our Coq compiler implement a verified topological sorting algorithm so that LoFIRRTL can be successfully generated.

In terms of efficiency, our OCaml FIRRTL compiler is generally faster (0.37 times on average) than {\sf firtool}. 
Only on 3 benchmarks from ChiselBook, our OCaml FIRRTL compiler used more compilation time than {\sf firtool}.
After an in-depth analysis, we found that this is attributed to our topological sorting algorithm based implementation of
the InferWidths transformation whose cost increase with the dependency graph.
For instance, on the program \texttt{DeepExecution} with 1220 lines,
 our OCaml FIRRTL compiler used 3 times more compilation time than {\sf firtool} and
 the Inferwidths pass used 30 times more execution time than {\sf firtool}.
   
  %We plan to implement a constraint solver as well and provide these two methods as an option.

