%!TEX root = main.tex

%\newcommand{\clock}{{\tt Clock}\xspace}
%\newcommand{\UInt}{{\tt UInt}\xspace}


A {\hifirrtl} program provides a high-level description of low-level RTL circuits. To a certain extent, it is %\emph{nondeterministic}
an under-specification in the sense that the bit-widths of its components are not fully determined.
The semantics we present is denotational, i.e., a {\hifirrtl} program
is interpreted as the set of (low-level) {\lofirrtl} programs that conform to it, where (1) all components are of ground types,
whose bit-widths are all explicitly given, (2) there are no \prettyll{when} statements, and (3) each component is connected at most once.
As a result, the semantics definition is essentially to specify which low-level RTL circuits conform to a given {\hifirrtl} program.

Furthermore, we can define an operational semantics of {\lofirrtl} programs, thus obtain a complete account of the formal semantics of {\firrtl}. However, as the operational semantics is not directly related to the formalization and verification of lowering transformations which is the main focus of this paper, they are relegated to Supplementary Material, Section~\ref{app:lofirrtl-op-sem}. 

Fix a {\hifirrtl} module $M$ which consists of a sequence of port declarations followed by statements. We shall represent a {\lofirrtl} module  %by a data structure called \emph{module graphs}. A module graph $G = (vm, ct)$ represents a module in {\hifirrtl} programs,
as a pair $(tm, cm)$ where $tm$ is a \emph{type mapping} that records the typing information of ground-type components in $M$, and $cm$ is a \emph{connection mapping} that records the connections between these components.
%
%A {\hifirrtl} module %$M$, \emph{the semantics of $M$ is defined as
%is interpreted as the set of such pairs that conform to it.
%
%Hence,
%The crux of the semantics definition is to
We define the conformance between $(tm, cm)$ and $M$ (denoted by $(tm, cm)\models M$)
as follows. Intuitively, %is defined as follows: \tl{to see how to improve this.}
\begin{itemize}
\item %At first, $vm$ is used to determine
we first utilize $tm$ to determine all the bit-widths of the components in $M$ and build a mapping $tmap$ that records the typing information of the (not necessarily ground-type) components in $M$. During the construction of $tmap$, we check that the types in $tm$ and $M$ are compatible and the component references in $M$ are well-scoped with respect to its \prettyll{when} statements,
	%
\item %to decide whether $(tm, cm) \models M$,
	we then utilize $tm$ and $tmap$ %together with the statements in $M$,
	to construct a connection mapping $cm'$.
	%moreover, during the construction of $(tm', cm')$,
\end{itemize}
We say that $(tm, cm) \models M$ if $tmap$ and $cm'$ are successfully constructed and $cm = cm'$.
%\fu{It seems that the above is a construction of $tmap$ and $cm'$ instead of the definition of conformance. When the construction succeeds, and $cm = cm'$, it
%yields a low level module that conforms to $M$. Can we first define what is conformance, then discuss how to compute all the low level modules that conform to $M$?
%Also, I suggest to move this conformance definition after definition of low-level circuits.}

The main technical difficulty %for defining $(tm, cm)\models M$
arises from the gap between $(tm, cm)$ and $M$: while $(tm, cm)$ is of low-level, $M$ is at high level which may contain aggregate types (i.e., vectors and bundles) as well as \prettyll{when} statements. As a result, %when checking $(tm, cm)\models M$,
it is necessary to %bridge this gap, in particular, to
find a match between the ground types in $tm$ and aggregate types in $M$, as well as a match between $cm$ and the connect statements in $M$, which requires ``expanding" the \prettyll{when} statements in $M$.
%
%In the sequel, we first define the concept of module graphs, then show how to construct $tmap$ from $vm$ and $M$, as well as how to construct a module graph from the empty graph by using $tmap$ and $M$.
%
%We fix ground types \clock and  \UInt{n} where $n$ is a positive integer.
%based on the discussion with zhilin, aggregate type is not needed.
%, and aggregate types consisting of vector type and bundle type. Typically, we write \UInt{n}.
%
%Let us fix a LoFIRRTL circuit $A$ that is free of \prettyll{inst of} statements. According to the syntax of LoFIRRTL, the top-level module of the circuit must be module $A$.
%
%A \emph{generalized connect statement} is a statement of the form $z <= e$ or $\mbox{\tt \prettyll{node}}\ z = e$, where $z$ and $e$ are called the \emph{head} and \emph{body} of the generalized connect statement, respectively. Intuitively, a generalized connect statement is either a connect statement $z <= e$ or a node-declaration statement $\mbox{\tt \prettyll{node}}\ z = e$.
%
%We denote by $\iports(A)$, $\oports(A)$, $\wires(A)$, $\nodes(A)$
%and $\registers(A)$ the sets of input ports, output ports, wires, nodes and registers of a module $A$, respectively.
%Furthermore, we use $\ports(A)$ to denote the set of ports $\iports(A) \uplus \oports(A)$,
%$\cpnts(A)$ to denote the set of components $\ports(A) \uplus \wires(A) \uplus \nodes(A) \uplus \registers(A)$,
%and $\ccpnts(A)$ to denote the set of components $\oports(A) \uplus \wires(A) \uplus \nodes(A)$ that must be connected to, % in the module $A$,
%where a component $z$ is connected to via a generalized connect statement whose head is $z$.
%Note that the input ports $\iports(A)$ are not connected to, moreover, the registers may not be connected to, thus they
%are not included in $\ccpnts(A)$. (If a register is not connected to, then it keeps the current value.)
Section~\ref{sec-construct-tmap} shows how to construct $tmap$ from $tm$ and $M$ and Section~\ref{sec-construct-graph} shows how to construct $cm'$ from $tmap$ and $M$. We start with some notations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\noindent\textbf{Expressions.}
$\exps$ (resp. $\refs$) denotes the set of {\hifirrtl} expressions (resp. references), that is, %the set of strings that can be
those derived  by the rules from $e$ (resp. $r$) in Figure~\ref{fig-firrtl-synt-simp}.
For convenience, we assume three special symbols $\svnone,\svundefined, \svinvalid \in \exps$.
%\tl{to check whether refs is used later.} \zhilin{it is used}

We use $\fieldseq$ to denote the set of field sequences, that is, $\vec{f} = f_1 \ldots f_k$ such that each $f_i$ is defined by the rule from $f$ in Figure~\ref{fig-firrtl-synt-simp}.
%For convenience, we denote
Each field $f_i$ is a triple $(x, b, t)$, where $x$ is the name of the field, $b$ is a Boolean value to denote whether the field is flipped or not, and $t$ is the type. Occasionally, an empty sequence is denoted by $\nil$.
%contains two special elements,   $$.

\noindent\textbf{Types.}
We denote by $\gtypes:= (\{\uint, \sint, \uintimp, \sintimp\} \times \Nat) \cup \{\clock,
\reset, \asyncreset\}$ the set of \emph{ground types}.
Note that $\uintimp$ and $\sintimp$ are introduced to denote the %explicit
ground types that are instantiated from the implicit ground types. %\tl{then why not use unitexp !?}
%
In particular, $\gtypes$ can model types of implicit width in $M$. For instance, $\uint$ where the width is implicit is captured by $(\uintimp, 0)$.
%\fu{The meaning of ``0'' is unclear here, does $i>0$ in $(\uintimp, i)$ denotes the width $i$ and $0$ denotes unknown?}
%\tl{so it is sugar?}
%\zhilin{UIntImp is used to check the compatibility between tm and tmap later on.}
%Similarly for $\sint$.
%Let $\bv=\{0,1\}^+$, i.e., the set of bit vectors.
%
%For a ground type $t = (\tau, n) \in \{UInt, SInt\} \times \Nat$, $n$ is the \emph{width} of $t$, denoted by $wd(t)$.
%

In $tm$,  the ground types are annotated with component sorts (e.g., input and output ports, wires, etc). To accommodate this, we define \emph{extended ground types} $\egtypes$ as %the set of vertex sorts.
\begin{itemize}
\item $(s, t)$ with $s \in \{\inport, \outport, \wire, \node\}$ and $t \in \gtypes$, or
\item $(\register, (t, clk, rsig, rval, b))$ such that $t \in \gtypes$, $clk, rsig, rval \in \exps$ and $b\in \{0,1\}$. (Intuitively, $clk$ represents a clock, $rsig$ represents a reset signal, and $rval$ represents a reset value, and $b$ is a Boolean value denoting whether the reset is asynchronous or not.)
\end{itemize}
We also assume $\svnone,\svundefined \in \egtypes$.

%Besides ground types,
The types for the components in $M$ (which may be of aggregate types) are referred to as \emph{FIRRTL types}, which are denoted by $\ftypes$ and  % to denote these types. More precisely, we define $\ftypes$ to be the set of elements of one of the following forms,
%Formally, $\ftypes$ is
defined inductively as follows.
\begin{itemize}
\item If $t \in \gtypes$, then $(\gtype, t) \in \ftypes$.
%
\item If $t \in \ftypes$ and $n \in \Nat$, then $(\vtype, t[n]) \in \ftypes$,
%
\item If $\vec{f} \in \fieldseq$, then $(\btype, \vec{f}) \in \ftypes$.
\end{itemize}
%Recall that the types where the bit widths are implicit are captured by $(\uintimp, 0)$ or $(\sintimp, 0)$.


\noindent\textbf{Identifiers.} For convenience, we assume an \emph{injective} identifier function $id$, which maps from the name of a component (e.g., wire, node, register) to a natural number; %Note that $id$ is an injective function.
$\range{id}$ denotes the range of $id$.
We shall use identifiers (instead of names) to refer to the components in $M$. On the other hand, %since $tm$ records the typing information for ground-type components, we define the
the identifiers in $tm$ are defined as $(v, m) \in \Nat \times \Nat$, which should be %aiming at linking them to
``linked" to the identifiers in $M$.
%Each component $cnt$ is associated with a pair of the form $(a, o) \in \Nat \times \Nat$, %each of which is corresponding to a component $cnt$ of $M$ with
%where $a = id(cnt)$ and
%$(a, o)$ is one of the following forms,
%each component $cnt$ of ground type in $M$ is associated with an identifier $(a, o)$, where
Intuitively, an identifier $(v, m)$ in $tm$ is used to refer to the ground-type components that are obtained by flattening the aggregate-type components in $M$. For an identifier $(v, m)$,
\begin{itemize}
\item either $v$ is the identifier of a ground-type component and $m = 0$,
%
\item or $v$ is the identifier of a vector-type component in $M$ and $m$ is an index in the vector,
%
\item or $v$ is the identifier of a bundle-type component in $M$ and $m$ is the offset of a ground-type field in the component.
\end{itemize}
Let $\gidset_M$ denote the set of such identifiers $(v,m)$.

The bundle-type component can be naturally represented as a tree where its (ground-type) fields are leaves. The offsets of these fields can be determined by a preorder traversal, which will be formally defined in Section~\ref{sec-construct-tmap}. As an example, for the bundle type
\begin{center}
\prettyll{output a: \{flip b: \{d: UInt<1>, e: UInt<2>\}, c: UInt<2>\}},
\end{center}
%
the ground-type fields are \prettyll{c}, \prettyll{d}, \prettyll{e}, and a preorder traversal gives an order $d, e, c$. As a result, the offsets of $d, e, c$ are $0$, $1$, and $2$ respectively. Note that %during the depth-first search,
the fields at the same level are visited %traversed
according to the order of occurrence in $M$. %Moreover, the field $b$ is omitted here since it is \emph{not} of ground type.
 	
\begin{definition}[Low-level RTL circuits]
A low-level RTL circuit is defined as a pair $(tm, cm)$, where $tm$ is a function from $\gidset_M$ to $\egtypes$
and $cm$ is function from $\gidset_M$ to $\exps$. %with a finite domain.
For convenience, $tm$ is called a type mapping and $cm$ is called a connection mapping.
\end{definition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Defining $tmap$ from $tm$ and $M$} \label{sec-construct-tmap}

\input{construct-tmap}

\subsection{Constructing a connection mapping from $tmap$ and $M$}\label{sec-construct-graph}

\input{construct-module-graph}
