The ``ExpandConnects'' pass expands the connections between aggregate
types into those between ground types. It also flattens the elements
in aggregate types to their corresponding ground types in declaration
statements. To expand the objects of aggregate types in a Firrtl
module, the process involves two phases. The first phase is to record
the types along with orientations and flipping information of the
objects, which helps determine the actual connection directions. For
example, if there is an input port of aggregate types with a field
marked as ``flip'', the ``flip'' mark will recursively apply to all
the atomic elements in that field. Since it is an input port, the
orientation is record as \source.  The second phase is to expand
the statements sequentially, and then produce a serial of new
statements.  For declaration statements, ground-typed elements are
generated corresponding to the original aggregate types and the
orientations. For connection statements, the connection directions are
determined by the orientations and flipping information from the first
phase. For subfields, there are two kinds of connections allowed for
HiFirrtl programs. One is connect subfields separately, in which case
the direction should already comply with the orientation of the
subfield elements. The other is to set connections between elements of
equivalent aggregate types. In such bundled connections, after
expansion process, individual subfield connections can be of different
directions.

At the beginning of a Firrtl module, ports are declared first. When
ports are defined with flipped subfields, the expansion process will
change those fields of input ports into output ports, and vice
versa. As a result, their orientations are changed between \source
and \sink.
 \begin{gather*}
   \label{expand-input}
   \infer[]
         {\seqq{{\tt expand\_connect} (\mbox{\prettyll{input v t}}) = pp}}
         {\begin{array}{c}
             \seqq{{\tt expand\_ref \;v} = [v_0, v_1,\dots v_n]}\\
             \seqq{ pp= {\tt map}\; (fun\; v_x \Rightarrow if\;{\tt is\_flipped}(v_x)\;then\; \mbox{\prettyll{output}} {\tt v_x}\;else\; \mbox{\prettyll{input}} {\tt v_x})\, [v_0, v_1,\dots v_n]}
           \end{array}
         }
 \end{gather*}
 The {\tt expand\_connect} function takes a statement as input and
produces a sequence of statements as output. In the case above, the
input statement is the port declaration ``\mbox{\prettyll{input v}}'',
and ``$pp$'' is the sequence of output. The ``$pp$'' is produced by
expanding the port {\tt v} to its atomic subfields with its type {\tt t},
and flipping the port orientation by judging whether the
subfields are marked with ``flip''. The {\tt map}
is the general operation for lists that applies a function over
lists. In the {\tt expand\_connect} function for
``\mbox{\prettyll{input v}}'', it applies the judging
{\tt is\_flipped} over all the expanded references of {\tt v}, then gives
the new sequence of declaration statements for input and output ports.

Elements such as wires and registers have duplex
orientation. Regardless of the flip subfields in the original
aggregate types, the resulting orientations will remain duplex. To
expand registers in aggregate types, we also need to expand the reset
values of the registers. Expanding wires is relatively simple and is
omitted here.
\begin{gather*}
  \label{expand-reg}
  \infer[]
        {\seqq{{\tt expand\_connect} (\mbox{\prettyll{reg v : t clk with : (reset => rst, rv)}}) = ss}}
        {\begin{array}{c}
            \seqq{{\tt expand\_ref \;}v = [v_0,\dots v_n]}\\
            \seqq{{\tt expand\_expr \;}rv = [rv_0,\dots rv_n]}\\
            \seqq{ s s= {\tt map}\; (fun\; v_x\; rv_x\; \Rightarrow \mbox{\prettyll{reg}}\; v_x:\,\mbox{\prettyll{t clk with:(reset=> rst,}}\,rv_x))}
            %% \rightline{
              \seqq{[v_0,\dots v_n], [rv_0,\dots rv_n]}
          \end{array}
        }
\end{gather*}
Here, {\tt expand\_connect} function takes
\mbox{\prettyll{reg v : t clk with : (reset => rst, rv)}} as input, which is the statement
to define a register {\tt v}. Similar to expanding a port, \xiaomu{add details} 

Nodes are also \source elements. They are only allowed to be declared
with passive aggregate types, which means no flip subfields. During
the expansion of node declarations, the passive type is checked;
otherwise, the result will be an empty sequence.
\begin{gather*}
  \label{expand-node}
  \infer[]
        {\seqq{{\tt expand\_connect} (\mbox{\prettyll{node v e}}) = ss}}
        {\begin{array}{c}
            \seqq{{\tt expand\_ref}\; v = [v_0, v_1,\dots v_n]}\\
            \seqq{{\tt expand\_expr}\; e = [e_0, e_1,\dots e_n]}\\
            \seqq{ s s= if\;{\tt is\_passive}(v_x)\; then\; {\tt map}\; (fun\; v_x\;e_x \Rightarrow \; \mbox{\prettyll{node}} {\tt v_x}\;{\tt e_x})\,[v_0, v_1,\dots v_n]\,[e_0, e_1,\dots e_n] else\; [::]}
          \end{array}
        } 
\end{gather*}

To expand \prettyll{v is invalid} statements, we
select only the \sink and \duplex elements to produce the sequence of
invalidations.
In HiFirrtl, it is allowed to have \source sub-elements in a bundle be
invalidated. The result of expansion will omit them.
\begin{gather*}
  \label{expand-invalid}
  \infer[]
        {\seqq{{\tt expand\_connect} (\mbox{\prettyll{v is invalid}}) = ss}}
        {\begin{array}{c}
            \seqq{{\tt expand\_ref \;}v = [v_0, v_1,\dots v_n]}\\
            \seqq{vs = {\tt filter}\; (fun\; v_x \Rightarrow orient (v_x) \neq source) [v_0, v_1,\dots v_n]}\\
            \seqq{ s s = {\tt map}\; (fun\; v_x \Rightarrow {\tt v_x}\;\mbox{\prettyll{is invalid}})\, vs}
          \end{array}
        } 
\end{gather*}

Connections between bundled elements contain a set of connections for
individual subfields. According to the indicated ``flip'' for a subfield, the
connection direction may differ from the original.
\begin{gather*}
   \label{expand-invalid}
   \infer[]
         {\seqq{{\tt expand\_connect} (\mbox{\prettyll{l <= r}}) = ss}}
         {\begin{array}{c}
             \seqq{{\tt expand\_ref }\; l = [l_0,\dots l_n]}\\
             \seqq{{\tt expand\_ref}\; r = [r_0,\dots r_n]}\\
             \seqq{ s s = {\tt map}\; (fun\; l_x\;r_x \Rightarrow if\;{\tt is\_flipped}(l_x)\; then\;\ {\tt r_x}\, \mbox{\prettyll{ <= }}\, {\tt l_x}\; else\;{\tt l_x}\, \mbox{\prettyll{ <= }}\, {\tt r_x})\,[l_0,\dots l_n]\,[r_0,\dots r_n]}
           \end{array}
         }
\end{gather*}
In the case above, the connection is set between two references of
elements {\tt l} and {\tt r}. When the connection takes an expression other than a
reference on the right-hand side, flip information is not taken into
account. The expansion simply expands the reference and the expression
and produces a new sequence of connections as follows.
\begin{gather*}
   \label{expand-invalid}
   \infer[]
         {\seqq{{\tt expand\_connect} (\mbox{\prettyll{l <= e}}) = ss}}
         {\begin{array}{c}
             \seqq{{\tt expand\_ref }\; l = [l_0,\dots l_n]}\\
             \seqq{{\tt expand\_expr}\; e = [e_0,\dots e_n]}\\
             \seqq{ s s = {\tt map}\; (fun\; l_x\;e_x \Rightarrow {\tt l_x}\, \mbox{\prettyll{ <= }}\, {\tt e_x})\,[l_0,\dots l_n]\,[e_0,\dots e_n]}
           \end{array}
         }
\end{gather*}

To summarize, a given Firrtl module can be expanded by applying the
{\tt expand\_connect} function to all its statements
sequentially. Then the module in return contains only the elements and
their connections in ground types.

