%!TEX root = main.tex

%In the sequel, we first introduce a concept of expression trees to specify the correctness property of {\tt expand\_branch}, then prove that {\tt expand\_branch} satisfies the property. 

\paragraph*{Specification}
As in the other phases, the correctness of {ExpandWhens} is specified by the formula:
%
\begin{equation}\label{eqn:expand-whens-spec}
    \forall tm, cm.\ (tm,\,cm) \models \, {\sf expandWhens}(M) \Rightarrow (tm,\,cm) \models \,M.
\end{equation}
%
This formula states that if the pair of the type mapping $tm$ and the connection mapping $cm$ conform to the program produced by {\sf expandWhens}$(M)$,
then the same pair of $tm$ and $cm$ should also conform to the original program $M$.
Note that this requires that an incorrect program is not translated to a conforming program
(for example, scoping errors like in Figure~\ref{fig:example}).
%Therefore, {\sf expandBranch} also checks for such scoping errors.

\paragraph*{Verification}
To prove this formula, we can assume that all components in $M$ have ground types,
and all connections are to ground-type components;
this simplifies creating the correct multiplexers.
%Let $(tm, cm) \models \expandwhens(M)$. We would like to show $(tm, cm) \models M$.
The construction of $tmap$ (which would record the aggregate types of components) becomes rather trivial;
also, the ports of $M$ and $tm$ are not supposed to change.
Therefore, {\sf expandWhens} mostly works on $cm$.
%Because $\expandwhens(M)$ does not affect $tm$. As a result, we focus on $cm$.
We first use the following lemma:
\begin{lemma}\label{lem:expand-whens-auxiliary}
$\connstmts(ecm)$ leads to a connect map $cm$ that contains $ecm$ as a submap,
i.e. if $ecm(x) \notin \{ \svnone, \svundefined \}$, then $ecm(x) = cm(x)$
for $cm_0 \xrightarrow[tm, tmap]{\connstmts(ecm)} cm$.
\end{lemma}
This is proven by induction over the elements of $ecm$,
using the property that $ecm$ cannot contain duplicate elements
(and therefore, $\connstmts(ecm)$ contains at most one connect statement to any given component).
Consequently, it is sufficient to prove that $\expandbranch(ss)$ constructs a map $ecm$
that contains the same assignments as the $cm$ constructed in the semantics $cm_0 \xrightarrow[tm, tmap]{ss} cm$.

The main inductive theorem is then essentially the following.
\begin{theorem}\label{thm:expand-whens-inductive}
If $\stmtstmap((tmap_0, scope_0), ss, tm) = (tmap, scope)$
and $\expandbranch(ss, \linebreak[0] ecm_0) = (ss_{\sf dcl}, ecm)$
and $cm_0 \xrightarrow[tm, tmap]{ss_{\sf dcl} ; \connstmts(ecm)} cm$,
then $cm_0 \xrightarrow[tm, tmap]{ss} cm$
(under some technical side conditions, e.g. that $ecm_0$ contains a connection for everything in scope according to $scope_0$).
\end{theorem}
This is proven by unfolding the constructions and checks in $\stmtstmap$, the semantics and $\expandbranch$ at the same pace.
We need to go through all the cases (e.g. in which branch a component is declared, in which branch(es) it is connected, whether it is still in scope),
leading to a long list of many cases to be handled.
We apply Lemma~\ref{lem:expand-whens-auxiliary} to unfold the semantic of $\connstmts$.
For every case, we show that the two mappings are (extensionally) equal.

Because of the recursive definition of $\stmtstmap$, the semantics and $\expandbranch$,
we use a mutual induction over statements and statement sequences.
The mutual induction also requires us to define the two Coq datatypes for statements and statement sequences mutually inductive;
we cannot just use the sequence or list types from the standard Coq library.

The complexity of the correctness proof lies mostly in the number of cases that need to be distinguished
and in the multitude of bookkeeping information from the side conditions that needs to be dragged through.

Using Theorem~\ref{thm:expand-whens-inductive},
it is then relatively easy to prove the specification \eqref{eqn:expand-whens-spec},
by filling the maps $tmap_0$, $scope_0$, $ecm_0$ and $cm_0$ with the information from the ports of module $M$.

%%%%%%%%%%%%%%%%%%% the original texts by David 
\hide{
The proof uses in a mutual induction over statements and statement sequences.
If a statement $s$ and a statement sequence $ss$ satisfy the correctness property, then also the statement sequence $s ; ss$.
Moreover, a statement \prettyll{when} $c$ \prettyll{:} $sst$ \prettyll{else} $ssf$
satisfies the correctness property if both $sst$ and $ssf$ satisfy it.
The mutual induction also requires us to define the two Coq datatypes for statements and statement sequences mutually inductive;
we cannot just use the sequence or list types from the standard Coq library.

The proof compares in detail the constructions and checks executed by {\sf stmtsTmap},
the construction of a connection mapping, and {\sf expandBranch}.
The main maps that should fit together are:
$scope$, $ecm$ (constructed by {\sf expandBranch}), $cm$, $tm$, and $tmap$ where $tmap$ is constructed from $tm$ by $\portsstmtstmap$.
Generally speaking,
the domain of every map in this list is a subdomain of the next map.
Also, the information in these maps needs to be consistent with each other, e.g. sources cannot be connected to.

We use these conditions (all types are ground-types, the five main maps fit together)
as background for the main inductive statement of the proof:
\begin{lemma}
If the constructions mentioned in Section 3 ({\sf stmtTmap}, the connection mapping, and {\expandbranch}) can be applied to statement sequence $ss$,
then 

;
\end{itemize}
Then:
\begin{itemize}
\item   an extensionally equal $cm$ is constructed by running $\xrightarrow[tm, tmap]{ss}$;
\item   the resulting maps fit together similarly;
\item   the connection map $cm'$ constructed by {\sf expandBranch} contains the same additions for $ss$
        as $cm$.
\end{itemize}
\end{lemma}
}
%%%%%%%%%%%%%%%%%%%%%%%

